---
title: "Integration"
format: html
editor: visual
---

## Preparation

### Set Up

Relevant libraries are loaded.

```{r, message = FALSE}
library(Seurat)
library(tidyverse)
library(stringr)
library(ggplot2)
library(patchwork)
library(SingleCellExperiment)
library(scater)
library(table1)
library(clustree)
library(openxlsx)
```

Define the path to get the data and save results.

```{r}
data_path <- "~/Dropbox/Special Project Single Cell/mouse/brain/data/seurat_data"

result_path <- "~/Dropbox/Special Project Single Cell/mouse/brain/results/integration"
```

### Data Load

The preprocessed data is loaded.

```{r}
seu_obj_preproc <- readRDS(paste(data_path,
                                 "seu_obj_preproc.rds",
                                 sep = "/"))
```

## Integration

### Clustering

The data is clustered using different resolutions by running the functions FindNeigbors and FindClusters.

```{r}
res <- c(0.1, 0.3, 0.5, 0.7)

seu_obj_preproc <- FindNeighbors(seu_obj_preproc, 
                                 dims = 1:30,
                                 verbose = FALSE)

seu_obj_preproc <- FindClusters(seu_obj_preproc, 
                                resolution = res,
                                verbose = FALSE)
```

### Batch Effects

There seems to be noticeable batch effects as cells cluster according to the timepoint in which they were taken.

```{r}
dimplot_pre <- DimPlot(seu_obj_preproc,
                       reduction = "umap", 
                       group.by = "experiment",
                       cols = c("#CC3333", "#0066CC"),
                       pt.size = 1,
                       alpha = 0.8) + 
  labs(title = "") + 
  xlab("UMAP1") +
  ylab("UMAP2")

ggsave(filename = "dimplot_pre.png",
       path = result_path,
       plot = dimplot_pre,
       width = 10,
       height = 6)

dimplot_pre
```

We color the cells based on both experiment and replicate.

```{r}
seu_obj_preproc$exp_rep <- paste(seu_obj_preproc$experiment,
                                 seu_obj_preproc$replicate,
                                 sep = " Rep ")
DimPlot(seu_obj_preproc, 
        reduction = "umap", 
        group.by = "exp_rep",
        cols = c("#990000", "#CC3333", "#FF9999", 
                 "#003399", "#0066CC", "#99CCFF"),
        pt.size = 1,
        alpha = 0.8) + 
  labs(title = "") + 
  xlab("UMAP1") +
  ylab("UMAP2")

ggsave(filename = "dimplot_pre_rep.png",
       path = result_path,
       width = 10,
       height = 6)
```

When coloring the cells according to their assigned cluster for a given resolution, it is clear, that all clusters are not present for a given experiment.

```{r}
list_dimplot <- list()

for (i in 1:length(res)){
  list_dimplot[[i]] <- DimPlot(seu_obj_preproc, 
                               reduction = "umap", 
                               group.by = paste("RNA_snn_res.", res[i], sep = ""),
                               split.by = "experiment",
                               cols = c("#CC3333", "#0066CC", "#339900", 
                                        "#FF9933", "#996633", "#FF3399"),
                               pt.size = 1,
                               alpha = 0.5) +
    labs(title = paste("Resolution", res[i])) + 
    xlab("UMAP1") +
    ylab("UMAP2") +
    theme(plot.title = element_text(hjust = 0))
  
  ggsave(filename = paste("dimplot_pre_",
                          res[i],
                          ".png",
                          sep = ""),
       path = result_path,
       plot = list_dimplot[[i]],
       width = 10,
       height = 6)
}

list_dimplot[[1]]
```

### Integration

We split the dataset into a list of 2 seurat objects (D0 and D11).

```{r}
seu_obj_split <- SplitObject(seu_obj_preproc, 
                             split.by = "experiment")
```

We normalize and identify variable features for each timepoint independently.

```{r}
for(i in 1:length(seu_obj_split)){
  seu_obj_split[[i]] <- NormalizeData(object = seu_obj_split[[i]],
                                      normalization.method = "LogNormalize",
                                      scale.factor = 10000,
                                      verbose = FALSE)
  
  seu_obj_split[[i]] <- FindVariableFeatures(object = seu_obj_split[[i]],
                                             selection.method = "vst",
                                             nfeatures = 2000,
                                             verbose = FALSE)
}
```

We select features that are repeatedly variable across datasets for integration.

```{r}
features <- SelectIntegrationFeatures(object.list = seu_obj_split)
```

The integration anchors are identified.

```{r}
anchors <- FindIntegrationAnchors(object.list = seu_obj_split,
                                  anchor.features = features,
                                  verbose = FALSE)
```

Finally, the data can be integrated using the anchors.

```{r}
seu_obj_integrated <- IntegrateData(anchorset = anchors,
                                    verbose = FALSE)
```

The integrated Seurat object is saved. 

```{r}
saveRDS(seu_obj_integrated, 
        file = "../data/seurat_data/seu_obj_integrated.rds")
```

We run the standard workflow for visualization and clustering.

```{r}
seu_obj_integrated <- ScaleData(seu_obj_integrated, 
                                verbose = FALSE)

seu_obj_integrated <- RunPCA(seu_obj_integrated,
                             verbose = FALSE)

seu_obj_integrated <- RunUMAP(seu_obj_integrated, 
                              reduction = "pca", 
                              dims = 1:30,
                              verbose = FALSE)

seu_obj_integrated <- FindNeighbors(seu_obj_integrated, 
                                    reduction = "pca", 
                                    dims = 1:30,
                                    verbose = FALSE)

seu_obj_integrated <- FindClusters(seu_obj_integrated, 
                                   resolution = res,
                                   verbose = FALSE)
```
We look at the results after integration.

```{r}
dimplot_post <- DimPlot(seu_obj_integrated,
                       reduction = "umap", 
                       group.by = "experiment",
                       cols = c("#CC3333", "#0066CC"),
                       pt.size = 1,
                       alpha = 0.8) + 
  labs(title = "") + 
  xlab("UMAP1") +
  ylab("UMAP2")

ggsave(filename = "dimplot_post.png",
       path = result_path,
       plot = dimplot_pre,
       width = 10,
       height = 6)

dimplot_post
```

We compare with the plot before integration.

```{r}
(dimplot_pre + theme(legend.position = "none") + labs(title = "Before integration")) | (dimplot_post + labs(title = "After integration"))

ggsave(filename = "dimplot_pre_post.png",
       path = result_path,
       width = 14,
       height = 6)
```

We see if the clusters for the integrated data are present for the two different timepoints.

```{r}
list_dimplot <- list()

for (i in 1:length(res)){
  list_dimplot[[i]] <- DimPlot(seu_obj_integrated, 
                               reduction = "umap", 
                               group.by = paste("integrated_snn_res.", res[i], sep = ""),
                               split.by = "experiment",
                               cols = c("#CC3333", "#0066CC", "#339900", 
                                        "#FF9933", "#996633", "#FF3399"),
                               pt.size = 1,
                               alpha = 0.5) +
    labs(title = paste("Resolution", res[i])) + 
    xlab("UMAP1") +
    ylab("UMAP2") +
    theme(plot.title = element_text(hjust = 0))
  
  ggsave(filename = paste("dimplot_post_",
                          res[i],
                          ".png",
                          sep = ""),
       path = result_path,
       plot = list_dimplot[[i]],
       width = 10,
       height = 6)
}

list_dimplot[[1]]
```

The figure above shows the results for resolution 0.1. The cells belonging to cluster 2 is mainly found for D11. We take a look at the clusters for all resolutions.

```{r}
list_dimplot <- list()

for (i in 1:length(res)){
  list_dimplot[[i]] <- DimPlot(seu_obj_integrated, 
                               reduction = "umap", 
                               group.by = paste("integrated_snn_res.", res[i], sep = ""),
                               cols = c("#CC3333", "#0066CC", "#339900", 
                                        "#FF9933", "#996633", "#FF3399"),
                               pt.size = 1,
                               alpha = 0.5) +
    labs(title = paste("Resolution", res[i])) + 
    xlab("UMAP1") +
    ylab("UMAP2") +
    theme(plot.title = element_text(hjust = 0))
}

wrap_plots(list_dimplot, 
           ncol = 2,
           nrow = 2,
           byrow = TRUE)

ggsave(filename = "clusters_post.png",
       path = result_path,
       width = 16,
       height = 10)
```

It is hard to decide on the resolution, as none of the plots seem perfect. Clustree can create a plot of a clustering tree showing the relationship between clusterings at different resolutions.

```{r}
clustree(seu_obj_integrated, 
         prefix = "integrated_snn_res.")
```

We want to resolutions which make samples to move into clusters dividing existing ones instead of moving them to other distant clusters. Thus, a resolution of 0.5 is chosen for now, but further investigation is necessary. 

```{r}
Idents(seu_obj_integrated) <- "integrated_snn_res.0.5"
```

## Cell Type assignment

We look at the expression of the different Vγ chains. 

```{r}
# Change assay to RNA.
DefaultAssay(seu_obj_integrated) <- "RNA"

# Extract gene names for gamma chains. 
genes <- sort(rownames(seu_obj_integrated))
gamma_chain <- genes[str_detect(genes, pattern = "Tcrg-V")]

list_featureplot <- FeaturePlot(seu_obj_integrated,
                                reduction = "umap",
                                features = gamma_chain,
                                min.cutoff = "q1",
                                label = TRUE,
                                combine = FALSE) 

list_featureplot <- lapply(X = list_featureplot, FUN = function(x){
  x + 
    xlab("UMAP1") +
    ylab("UMAP2")
})

wrap_plots(list_featureplot, 
           ncol = 2,
           nrow = 3,
           byrow = TRUE)

ggsave(filename = "featureplot_Vg.png",
       path = result_path,
       width = 10,
       height = 12)
```

Cluster 0 express Vγ6 while cluster 2 express Vγ4. We take a look at the expression of characteristic interleukins. We see which genes stand out in the clusters in compared to the others. 

```{r}
# Join layers.
seu_obj_integrated[["RNA"]] <- JoinLayers(seu_obj_integrated[["RNA"]])

# List to collects the markers for each cluster.
list_clusters <- list()

# Find markers.
for (i in 1:length(table(Idents(seu_obj_integrated)))){
  
  list_clusters[[i]] <- data.frame(FindMarkers(seu_obj_integrated,
                                               ident.1 = i-1,
                                               only.pos = TRUE,
                                               min.pct = 0.25,
                                               min.diff.pct = -Inf,
                                               logfc.threshold = 1,
                                               verbose = FALSE))
}

# Save the results in excel.
wb <- createWorkbook()
 
for (i in 1:length(list_clusters)) {
  addWorksheet(wb, 
               sheetName = paste("cluster_", 
                                 i-1,
                                 sep = ""))
  writeData(wb, 
            sheet = i, 
            x = list_clusters[[i]], 
            rowNames = TRUE)
}
 
saveWorkbook(wb, paste(result_path,
                       "cluster_markers.xlsx", 
                       sep = "/"))
```

We see if the clusters express genes typical for type 1 or type 3 γδT cells.

```{r}
gdt_type_1 <- c("Tbx21", "Ifng", "Gzma", 
                "Gzmb", "Stat1", "Irf1",
                "Cd27")

gdt_type_3 <- c("Rora", "Rorc", "Il22", 
                "Il17a", "Il17f", "Il23r", 
                "Ccr6", "Cd44", "Csf2")

for (i in 1:length(list_clusters)){
  print(paste("Cluster", i-1))
  print(paste("Number of conserved markers for type 1 signature:",
              sum(rownames(list_clusters[[i]]) %in% gdt_type_1)))
  print(paste("Number of conserved markers for type 3 signature:",
              sum(rownames(list_clusters[[i]]) %in% gdt_type_3)))
}
```

It looks like cluster 0 and 2 might consist of type 3 γδT cells, while cluster 1 and 4 express genes from the type 1 signature. Cluster 3 seems express genes from type 1.

```{r}
list_featureplot <- FeaturePlot(seu_obj_integrated,
                                reduction = "umap",
                                features = gdt_type_1,
                                min.cutoff = "q1",
                                label = TRUE,
                                combine = FALSE) 

list_featureplot <- lapply(X = list_featureplot, FUN = function(x){
  x + 
    xlab("UMAP1") +
    ylab("UMAP2")
})

wrap_plots(list_featureplot, 
           ncol = 3,
           nrow = 3,
           byrow = TRUE)

ggsave(filename = "featureplot_type1_sig.png",
       path = result_path,
       width = 14,
       height = 12)
```

We do the same for the type 3 signature.

```{r}
list_featureplot <- FeaturePlot(seu_obj_integrated,
                                reduction = "umap",
                                features = gdt_type_3,
                                min.cutoff = "q1",
                                label = TRUE,
                                combine = FALSE) 

list_featureplot <- lapply(X = list_featureplot, FUN = function(x){
  x + 
    xlab("UMAP1") +
    ylab("UMAP2")
})

wrap_plots(list_featureplot, 
           ncol = 3,
           nrow = 3,
           byrow = TRUE)

ggsave(filename = "featureplot_type3_sig.png",
       path = result_path,
       width = 14,
       height = 12)
```

Cluster 3 seems express both genes from the type 1 and the type 3 signature. The clusters are annotated. 

```{r}
seu_obj_annotated <- RenameIdents(object = seu_obj_integrated,
                                  "0" = "Type 3",
                                  "1" = "Type 1",
                                  "2" = "Type 3",
                                  "3" = "UAN",
                                  "4" = "Type 1")

seu_obj_annotated$cell_type <- factor(Idents(seu_obj_annotated),
                                      levels = c("Type 1", "Type 3", "UAN"))

DimPlot(seu_obj_annotated,
        reduction = "umap",
        group.by = "cell_type",
        cols = c("#CC3333", "#0066CC", "#339900",
                 "#FF9933", "#996633", "#FF3399"),
        pt.size = 1,
        alpha = 0.5) +
  labs(title = "Resolution 0.5") + 
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme(plot.title = element_text(hjust = 0))

ggsave(filename = "cell_annotation.png",
       path = result_path,
       width = 10,
       height = 6)
```

We look at the plot for each experiment. The number of type 3 cells seems to increase significantly.

```{r}
DimPlot(seu_obj_annotated,
        reduction = "umap",
        group.by = "cell_type",
        split.by = "experiment",
        cols = c("#CC3333", "#0066CC", "#339900",
                 "#FF9933", "#996633", "#FF3399"),
        pt.size = 1,
        alpha = 0.5) +
  labs(title = "Resolution 0.5") + 
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme(plot.title = element_text(hjust = 0))

ggsave(filename = "cell_annotation_exp.png",
       path = result_path,
       width = 10,
       height = 6)
```

The annotated Seurat object is saved. 

```{r}
saveRDS(seu_obj_annotated, 
        file = "../data/seurat_data/seu_obj_annotated.rds")
```

## This is just mess

```{r}
DimPlot(seu_obj_integrated,
        reduction = "umap",
        group.by = "integrated_snn_res.0.5",
        cols = c("#CC3333", "#0066CC", "#339900",
                 "#FF9933", "#996633", "#FF3399"),
        pt.size = 1,
        alpha = 0.5) +
  labs(title = "Resolution 0.5") + 
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme(plot.title = element_text(hjust = 0)) + 
  geom_vline(xintercept = c(5.5, 7.7)) +
  geom_hline(yintercept = c(1.5, 3))
```

```{r}
seu_obj_integrated@meta.data$umap1 <- seu_obj_integrated@reductions$umap@cell.embeddings[, 1]

seu_obj_integrated@meta.data$umap2 <- seu_obj_integrated@reductions$umap@cell.embeddings[, 2]

seu_obj_sub <- subset(seu_obj_integrated,
                      subset = umap1 > 5.5 & umap1 < 7.7 & umap2 > 1.5 & umap2 < 3)

```
